Combinator Parsers: a Monad Transformer Approach

### 2013-05-08 article.tex
massively simplify grammar, in order to allow examples to be much simpler and more focused on the important issues of parser combinators and monad transformers:
 - remove string, number tokens
 - simplify symbol definition

create new simple parser implementation, and describe it in write-up, as simplest version 1 parse tree and recognization demonstration

Form         :=   Special  |  Application  |  Symbol  |  
                  String   |  Number

@@

Define       :=  'define'  String(?)  Symbol  Form

@@

Number       :=  \d(+)

Symbol       :=  ( \w  |  Schar)  ( \w  |  \d  |  Schar)(*)

Schar        :=  (oneof "<>!@#$%^&*_-+=|:?")

String       :=  '"'  ( Escape  |  (not  ( '"'  |  '\')))(*)  '"'

Escape       :=  '\'  ( '"'  |  '\' )

@@

Comment      :=  ';'  (not '\n')(*)
\end{verbatim}
With the additional rule that whitespace and comments 
may appear in any amount before any token.
Tokens are:  {, }, (, ), Symbol, String, and Number.

@@

Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ANumber Integer
    | ASymbol String
    | AString String
    | ALambda [String] [AST]
    | ADefine (Maybe String) String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}
Plus we'll need a type for our parsers.  We'll use the basic type mentioned 
above -- [t] -> m (a, [t]) -- as our starting point.  As luck would have it, that
corresponds to our first monad transformer:  StateT (Control.Monad.State), so we
can write:
\begin{verbatim}
type Parser a = StateT [Char] Maybe a
\end{verbatim}
Which means that our Parser is a State monad transformer, with the Maybe type
as its underlying Monad parameter.

Now let's get down to the business of building parsers, starting with parsers
for our most basic syntactic elements:  tokens.
These include the four braces, symbols, strings, numbers, whitespace, and comments.
To recognize braces, we need to re-implement the item and check combinators that
we did before.  Fortunately, using the StateT transformer makes it simpler
to do so:
\begin{verbatim}
item :: (MonadState [t] m, Plus m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> zero;
\end{verbatim}
Although we now also need to discuss the Plus type class, why we need it instead of
Alternative, and what its instances are.

\section{Example 2: error-reporting}

>>>>

Form         :=   Symbol  |  Special  |  Application

Symbol       :=  [a-zA-Z](+)

@@

Define       :=  'define'  Symbol  Form

@@

@@

Comment      :=  ';'  (not '\n')(*)
\end{verbatim}
With the additional rule that whitespace and comments may appear in any 
amount before any token.  Tokens are:  {, }, (, ), and Symbol.

@@

Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ANumber Integer
    | ASymbol String
    | AString String
    | ALambda [String] [AST]
    | ADefine String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}
Plus we'll need a type for our parsers.  We'll use the basic type mentioned 
above -- \begin{verbatim}[t] -> m (a, [t]) \end{verbatim}
 -- as our starting point.  This type corresponds to our first monad transformer:  
 StateT (from Control.Monad.State), so we can write:
\begin{verbatim}
type Parser a = StateT [Char] Maybe a
\end{verbatim}
and a function for running a parser:
\begin{verbatim}
runParser :: Parser a -> String -> Maybe (a, String)
runParser = runStateT
\end{verbatim}
Which means that our Parser is a State monad transformer, with the Maybe type
as its underlying Monad parameter.

The signficance of the Maybe parameter is two-fold:  it allows computational
effects to express first success versus failure, using the Just and Nothing 
constructors, respectively, and second backtracking/alternation.  We'll
capture both of these characteristics with using a type class, Plus:
\begin{verbatim}
class Applicative f => Plus f where
  zero :: f a
  (<+>) :: f a -> f a -> f a
\end{verbatim}
This type class is identical in intent to the Alternative type class from
Control.Applicative, with zero corresponding to empty and <+> to \verb+<|>+; 
however, we aren't able to use Alternative directly because
other of the monad transformers have unsuitable instances for our purposes,
and as we'll see later, inability to swap out such instances is one of the 
limitations of the monad transformer approach.

To get around this limitation, we'll create instances for the Plus type class:
\begin{verbatim}
instance Plus Maybe where
  zero = Nothing
  Nothing <+> y = y
  x       <+> _ = x

instance (Monad m, Plus m) => Plus (StateT s m) where
  zero = StateT (const zero)
  StateT f <+> StateT g = StateT (\s -> f s <+> g s)
\end{verbatim}
Note that the StateT instance relies on its monad parameter, m, also being a 
member of the Plus type class.  This works out for our parser, since m is Maybe
and we do have a Plus instance for Maybe.

\subsection{Token parsers}
Now let's get down to the business of building parsers, starting with parsers
for our most basic syntactic elements:  tokens.
These include the four braces, symbols, strings, numbers, whitespace, and comments.
To recognize braces, we need to re-implement the item and check combinators that
we did before.  Using the MonadState and Plus type classes, 
we can now implement the basic item parser as a stateful
computation that fails if the state is empty:
\begin{verbatim}
item :: (MonadState [t] m, Plus m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> zero;
\end{verbatim}
And we can easily implement a new version of the check combinator:
\begin{verbatim}
check :: (Monad m, Plus m) => (a -> Bool) -> m a -> m a
check f p =
    p >>= \x ->
    if (f x) then return x else zero
\end{verbatim}
And use both of them to build a new combinator, satisfy, that checks whether
the next character meets a condition:
\begin{verbatim}
satisfy :: (MonadState [t] m, Plus m) => (t -> Bool) -> m t
satisfy = flip check item
\end{verbatim}
And finally a parser that checks that the next character is equal to a
given character:
\begin{verbatim}
literal :: (Eq t, MonadState [t] m, Plus m) => t -> m t
literal c = satisfy ((==) c)
\end{verbatim}
Now we're ready to build parsers for the four bracket tokens.  
An exercise for the reader:  what is the inferred type of each parser?
Let's first build a parser for open curly brackets:
\begin{verbatim}
ocurly = literal '{'
\end{verbatim}
And let's try it out in ghci to make sure it does what we expect:
\begin{verbatim}
ghci$ runParser ocurly "{abc"
Just ('{', "abc")

ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
[Nothing, Nothing, Nothing]
\end{verbatim}
Good, it correctly accepts open curly brackets and rejects everything else.
But what about the spec that said that comments and whitespace can occur 
before any token -- will our parser be able to handle that?
\begin{verbatim}
ghci$ map (runParser  ocurly) ["}abc", "(abc", "abc"]
Nothing
\end{verbatim}
Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a combinator (which we'll call munch) that takes in a parser, 
throws away all leading junk, then runs the parser:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')

munch p = many0 (whitespace <+> comment) *> p
\end{verbatim}
The comment parser uses a new combinator, not1, which is built out of a new type class,
Switch, whose intention is to model computations which can be switched from successful
to failing and vice versa:
\begin{verbatim}
class Switch f where
  switch :: f a -> f ()

instance Switch Maybe where
  switch (Just _) = Nothing
  switch Nothing  = Just ()

instance (Functor m, Switch m) => Switch (StateT s m) where
  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)

not1 :: (MonadState [t] m, Plus m, Switch m) => m a -> m t
not1 p = switch p *> item
\end{verbatim}
And now let's fix the ocurly parser using munch and try it out:
\begin{verbatim}
ocurly = munch $ literal '{'

ghci$ map (runParser ocurly) [";abc\n  {def", "{def", "(def", "}def"]
[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
\end{verbatim}
Excellent!  Now the open curly parser correctly throws away whitespace and comments,
and succeeds, consuming one character, when the first non-junk character is an open 
curly brace, and fails otherwise.
We can now define parsers for the other three brace tokens:
\begin{verbatim}
ccurly = munch $ literal '}'
oparen = munch $ literal '('
cparen = munch $ literal ')'
\end{verbatim}
The parser for our last token, Symbol, is slightly more complicated because it allows
multiple characters and multiple lengths.  We'll use the satisfy combinator along with 
a function that checks whether a character is one of the legal symbol characters:
\begin{verbatim}
symbol = munch $ many1 char
  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))
\end{verbatim}
Note the use of the many1 combinator to introduce repetition of any length of at least one,
and we're again using the munch combinator to throw away leading junk.

\subsection{Syntactic structures}
Whereas our token parsers dealt with the syntactic primitives of the Woof grammar, 
the remaining parsers will have to implement grammar rules that combine the tokens
into syntactic structures.
The first combining form is a function application.  The rule says that it's delimited
by matching parentheses, in between which must appear one form as the operator, followed
by any number of forms as the arguments to which the operator is applied.  We can
say that like so:
\begin{verbatim}
application =
    oparen      >>
    form        >>= \op ->
    many0 form  >>= \args ->
    cparen      >>
    return (AApp op args)
\end{verbatim}
Although this implementation uses the monadic operators >> and >>=, it could
equivalently have been implemented using the applicative combinators *> and <*>:
\begin{verbatim}
application' =
    oparen        *>
    pure AApp    <*>
    form         <*>
    many0 form   <*
    cparen
\end{verbatim}
Next, we move on to define, which is also a straightforward translation from the grammar:
\begin{verbatim}
define =
    ocurly                       *>
    check (== "define") symbol   *>
    pure ADefine                <*>
    symbol                      <*>
    form                        <*
    ccurly
\end{verbatim}
Lambda is more complicated -- not only does it have additional syntax with its parameter list,
but we also need to ensure that the parameter names are unique.  We can do that using the 
check combinator again:
\begin{verbatim}
lambda =
    ocurly                          >>
    check (== "lambda") symbol      >>
    ocurly                          >>
    check distinct (many0 symbol)   >>= \params ->
    ccurly                          >>
    many1 form                      >>= \bodies ->
    ccurly                          >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed, and parsing will
continue; if there's a repeated symbol, the parsing will stop with a failure.

\subsection{Finishing touches}
Now all we have left to do is to define the special, form (which we've already used to build
the previous parsers), and woof parsers.  The special parser is simple:  it parses either 
define or lambda:
\begin{verbatim}
special = define <+> lambda
\end{verbatim}
The form parser parsers either a symbol, application, or a special form.  We could just write:
\begin{verbatim}
form = symbol <+> application <+> special
\end{verbatim}
except that the types don't match -- symbol's type parameter is a String, whereas the other two
have ASTs, so we can just fix that by mapping the correct function over symbol:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}
The final parser, woof, needs to not only parse all the forms, but also make sure that the
entire input is consumed.  The end of input check is done with the end parser:
\begin{verbatim}
end = switch item
\end{verbatim}
And we also need to munch any trailing comments and whitespace, so our final woof parser is:
\begin{verbatim}
woof = many1 form <* munch end
\end{verbatim}

\subsection{Examples}
Let's look at a few examples of our parser in action:

\section{Example 2: error-reporting}
While our first attempt at a parser was useful, when it encountered malformed input, it wasn't
helpful -- all it could do was throw up its hands in despair and give up.  However, we'd like
an error message that tells us what the problem was and where it was encountered whenever bad
input is encountered.  Otherwise, we'd waste time finding the problem by eye -- not a lot of 
fun.  Here are some examples of bad input that we'd like to get specific error messages for,
but currently can't:
\begin{verbatim}
unrecognized character:
application missing operator:
application missing close parentheses:
special form missing close curly:
invalid special form name:
duplicate lambda parameters:
lambda missing body forms:
lambda missing parameter list:
define missing symbol:
define missing form:
\end{verbatim}
